Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-28 forecastflows.tex
integrate JD comments

We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation % jregetz16april: Maybe “data simulation” needs to appear in the figure? Or wait to mention simulation until referencing Fig 2? [FLAG: what do you think? I agree you should choose on of the two options]
(Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, fig. \ref{fig:workflow}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This step should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error. % vvdm16april: to answer Jonathan comment about verbal model
From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, fig. \ref{fig:workflow}). The general idea is to start with a relatively simple model that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterizations).

@@

\columnbreak
\vspace*{1mm}
Blabla... \\
% We say here briefly how great the new approach is or something.
A universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}. Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge. And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.

@@

We believe our workflow could help advance ecological science and its applications, but widespread use of it requires overcoming major hurdles that pervade science.  
% One well-known hurdle is the pressure to publish, which can lower research standards and make the added effort of this workflow seem ill-placed. This may be especially true for those who see science rewarded mostly through the shear number of publications. But for those more focused on the long-term value of their work---for example, how well cited their papers are over a longer-time scale---we think this workflow can help. % jdavies15april: I am not sure this needs to be said, but could mention good model fitting takes time and is an iterative and learning process.
The first is pressure to publish quickly, which can be at odds with the reality that good model fitting is inherently iterative and takes time. The second is reluctance to embrace adoption of open science practices that ensure modeling efforts are fully transparent and reproducible. It is imperative that we continue to place increasing value on research that is carefully developed, openly collaborative (including both data and code), and transparent about areas of uncertainty. Especially in ecology where samples sizes are low and effects are likely non-linear and complex, we believe adopting the workflow will pay off in the longer term, both for individual researchers focused on making enduring contributions, and for the community as a whole focused on improving future data collection and model development (see Fig).

@@

Advancing ecology to where most researchers use models built more flexibly from ecological theory and insights applied to their ecological systems will not happen rapidly without a major shift in training. Much of ecology still divides the world into training for those focused on several groups: those who gather data and learn a limited set of pre-built models versus those who develop more complex models. In ecological training today, researchers who conduct field and lab studies often learn a highly limited set of particular statistical tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test).

Instead, ecologists trained in a limited set of tests are expected to collaborate with others when they need more complex models, who were trained more in model development (though often for highly specific applications, such as wildlife population estimates, where they may rarely develop entirely new generative models). These two groups further differ from process-based modelers, who often train in physical and ecophysiological processes and how to abstract them into mainly deterministic models. % process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms 
Few of these groups have integrated data simulation into their statistical or scientific workflows. Simulation is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data. While specialization is valuable, we argue the fundamental % training
separation in ecology has overly-siloed these groups and prevented more rapid progress.

Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in the workflow would focus on learning to generate questions and then models, and then to simulate data from them. This would mean training all ecologists to link the ecological processes they study with the mathematical models that may describe them. Through this and the use retrodictive checks, most ecologists would be better equipped to think through what parameters are most critical to their question and/or aim (e.g., management), and also gain a much stronger connection to the level of uncertainty in many of ecological estimates. Empirical ecologists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models. Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may suddenly see bridges from their work to empirical data and forecasting.
% Perhaps the only drawback would be to those focused on statistical models, whose gate-keeping would suddenly fail them and they would have to do more that just say how amazing they are... 
Those focused on learning complex statistical models may find many others in the field now share their excitement and interest in more generative models, and could focus on adapting approaches for better parameter estimation and improved forecasts.

@@

% jdavies15april: This could be the Concluding section highlighting the general advances provided by adopting your framework (and perhaps a better place for text currently at end of section 3.2).
Today model development in ecology is rarely transparent, which limits how easily the research community can understand modes, and thus identify potential issues. Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations. However, we believe our workflow provides an tractable step to fixing this. By focusing on model development more tightly tied to ecological expertise, we argue this workflow should broaden the community that contributes to model development. As ecologists are increasingly expanding their computational toolkits, many field, and lab, and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological domains of interest, and---most importantly---to interrogate those models.

>>>>

We argue a workflow that moves along the data-model space in a coherent sequence of steps % with repeated data simulation % jregetz16april: Maybe “data simulation” needs to appear in the figure? Or wait to mention simulation until referencing Fig 2? [FLAG: what do you think? I agree you should choose on of the two options]
(Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (step 1, fig. \ref{fig:workflow}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This step should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error. % vvdm16april: to answer Jonathan comment about verbal model
From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question (step 2, fig. \ref{fig:workflow}). The general idea is to start with a relatively simple model that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires considering alternative parameterizations).

@@

\columnbreak
\vspace*{1mm}
Blabla... \\
% We say here briefly how great the new approach is or something.
A universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}. Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge. And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community, potentially improving trend estimations.

@@

We believe our workflow could help advance ecological science and its applications, but widespread use of it requires overcoming major hurdles that pervade science.  
% One well-known hurdle is the pressure to publish, which can lower research standards and make the added effort of this workflow seem ill-placed. This may be especially true for those who see science rewarded mostly through the shear number of publications. But for those more focused on the long-term value of their work---for example, how well cited their papers are over a longer-time scale---we think this workflow can help. % jdavies15april: I am not sure this needs to be said, but could mention good model fitting takes time and is an iterative and learning process.
The first is pressure to publish quickly, which can be at odds with the reality that good model fitting is inherently iterative and takes time. The second is reluctance to embrace adoption of open science practices that ensure modeling efforts are fully transparent and reproducible. It is imperative that we continue to place increasing value on research that is carefully developed, openly collaborative (including both data and code), and transparent about areas of uncertainty. Especially in ecology where samples sizes are low and effects are likely non-linear and complex, we believe adopting a workflow such as the one we propose here will pay off in the longer term, both for individual researchers focused on making enduring contributions, and for the community as a whole focused on improving future data collection and model development (see Fig).

@@

Advancing ecology to where most researchers use models built more flexibly from ecological theory and insights applied to their ecological systems will not happen rapidly without a major shift in training. Much of ecology still divides the world into training for those who gather data and learn a limited set of pre-built models versus those who develop more complex models. In ecological training today, researchers who conduct field and lab studies often learn a highly limited set of particular statistical tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test).

When ecologists trained in a limited set of tests require more complex models, they are expected to collaborate with others trained more in model development (though often for highly specific applications, such as wildlife population estimates, where generative models are often predefined). These two groups further differ from process-based modelers, who often train in physical and ecophysiological processes and how to abstract them into mainly deterministic models. % process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms 
Neither group has fully integrated data simulation into their statistical or scientific workflows. Simulation is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data. While specialization is valuable, we argue the fundamental % training
separation in ecology has overly-siloed these groups and prevented more rapid progress.

Instead of siloing ecological training into experimental design and tests that may match certain designs (though rarely do for ecological data, CITES) versus model design and development, we suggest unified training in the workflow would focus on learning to generate questions and then models, and then to simulate data from them. This would mean training to link the ecological processes they study with the mathematical models that may describe them. Through this and the use retrodictive checks, most ecologists would be better equipped to think through what parameters are most critical to their question and/or aim (e.g., management), and also gain a much stronger connection to the level of uncertainty in many of ecological estimates. Empiricists % change suggested by JD, was empirical ecologists before...
 would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models. Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may suddenly see bridges from their work to empirical data and forecasting.
% Perhaps the only drawback would be to those focused on statistical models, whose gate-keeping would suddenly fail them and they would have to do more that just say how amazing they are... 
Those focused on learning complex statistical models may find many others in the field now share their excitement and interest in more generative models, and could focus on adapting approaches for better parameter estimation and improved forecasts.

@@

% jdavies15april: This could be the Concluding section highlighting the general advances provided by adopting your framework (and perhaps a better place for text currently at end of section 3.2).
Today model development in ecology is rarely transparent, which limits how easily the research community can understand models, and thus identify potential issues. Instead of broad inclusive conversations about how to improve models to advance our ecological understanding, a significant portion of scientific debate today has become lost in methodological considerations. However, we believe our workflow provides a tractable step to fixing this. By focusing on model development more tightly tied to ecological expertise, we argue this workflow should broaden the community that contributes to model development. It may also help resolve apparent conflicts by identifying where divergences in model predictions emerge, whether from differences in assumptions, model structures, or other aspects of the modeling process.
As ecologists are increasingly expanding their computational toolkits, many field, and lab, and other forms of `empirical' ecologists have the basic tools to follow this workflow to build models that better represent their ecological domains of interest, and---most importantly---to interrogate those models.
